FAT Pointer based range addresses

### 2024-09-05 papers/EuroSys/Paper/paper.tex
latest changes

%%
%% This is file `sample-acmsmall-submission.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall-submission')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall-submission.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall,screen,review]{acmart}

% \usepackage[style=authoryear, useprefix = false]{biblatex}

@@

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
% \setcopyright{acmlicensed}
% \copyrightyear{2018}
% \acmYear{2018}
% \acmDOI{XXXXXXX.XXXXXXX}

%%
%% These commands are for a JOURNAL article.
% \acmJournal{JACM}
% \acmVolume{37}
% \acmNumber{4}
% \acmArticle{111}
% \acmMonth{8}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{FAT-Pointer based range addresses}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Akilan Selvacoumar}
% \authornote{Both authors contributed equally to this research.}
\email{as251@hw.ac.uk}
\orcid{1234-5678-9012}
% \author{G.K.M. Tobin}
% \authornotemark[1]
% \email{webmaster@marysville-ohio.com}
\affiliation{%
  \institution{Heriot Watt University}
  % \streetaddress{P.O. Box 1212}
  % \city{Dublin}
  % \state{Ohio}
  \country{UK}
  % \postcode{43017-6221}
}

% \author{Lars Th{\o}rv{\"a}ld}
% \affiliation{%
%   \institution{The Th{\o}rv{\"a}ld Group}
%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
%   \city{Hekla}
%   \country{Iceland}}
% \email{larst@affiliation.org}

@@

% \author{Aparna Patel}
% \affiliation{%
%  \institution{Rajiv Gandhi University}
%  \streetaddress{Rono-Hills}
%  \city{Doimukh}
%  \state{Arunachal Pradesh}
%  \country{India}}

% \author{Huifen Chan}
% \affiliation{%
%   \institution{Tsinghua University}
%   \streetaddress{30 Shuangqing Rd}
%   \city{Haidian Qu}
%   \state{Beijing Shi}
%   \country{China}}

% \author{Charles Palmer}
% \affiliation{%
%   \institution{Palmer Research Laboratories}
%   \streetaddress{8600 Datapoint Drive}
%   \city{San Antonio}
%   \state{Texas}
%   \country{USA}
%   \postcode{78229}}
% \email{cpalmer@prl.com}

% \author{John Smith}
% \affiliation{%
%   \institution{The Th{\o}rv{\"a}ld Group}
%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
%   \city{Hekla}
%   \country{Iceland}}
% \email{jsmith@affiliation.org}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  % Draft 1
  % The gap between application workloads and capacity of TLB is on the 
  % rise. Previous work exploits this by physically contiguous 
  % memory. Capability-based addressing in the CHERI architecture is 
  % designed to improve hardware-level system security.
  % These security mechanisms designed in place can also 
  % behave as accelerators to standard user-space memory allocators. 
  % The property of CHERI which is to use FAT-Pointers to store
  % the upper and lower bounds can also be represented as ranges of 
  % memory addresses which are physically contiguous. 
  % <Talking about the results>.
  % Draft 2
  The increasing disparity between application workloads and the capacity of Translation Lookaside Buffers (TLB) 
  has prompted researchers to explore innovative solutions to mitigate this gap. One such approach involves 
  leveraging physically contiguous memory to optimize TLB utilization. Concurrently, advancements in hardware-level 
  system security, exemplified by the Capability Hardware Enhanced RISC Instructions (CHERI) architecture, offer 
  additional opportunities for improving memory management and security.

@@

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
% \begin{CCSXML}
% <ccs2012>
%  <concept>
%   <concept_id>00000000.0000000.0000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>500</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>00000000.00000000.00000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>300</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>00000000.00000000.00000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>00000000.00000000.00000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
% </ccs2012>
% \end{CCSXML}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
% \keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
%   Your, Paper}

@@

% Sample citation
% \cite{ad-wood-2003}

@@

\section{Introduction}
In the dynamic landscape of computing, the pursuit of optimal performance is a constant endeavor, 
especially as applications evolve to handle increasingly complex workloads. 
One critical aspect influencing performance is memory management, where efficient 
utilization of resources is paramount. Translation Lookaside Buffers (TLBs) play a 
pivotal role in this regard, expediting memory access by storing recently accessed memory translations.
However, as applications grow in size and complexity, the capacity of TLBs often struggles to 
keep pace, leading to performance bottlenecks. To address this challenge, researchers have 
turned to innovative solutions, one of which involves harnessing the benefits of huge pages.
Huge pages, also known as large pages, allow for the allocation of memory in significantly 
larger chunks compared to traditional small pages. By reducing the number of TLB entries 
needed to access a given amount of memory, huge pages offer a potential avenue for optimizing 
TLB utilization and thereby enhancing overall system performance.

@@

In this context, the integration of huge pages into memory management 
strategies alongside capability-based addressing in architectures like 
CHERI offers a compelling synergy. By optimizing TLB utilization through the 
utilization of huge pages and leveraging the security features of capability-based addressing, 
significant performance improvements can be realized. This approach not only enhances 
system security but also accelerates memory access.

% TODO: 
% - Add references for FlexPointer, Range Memory Mapping (RMM), Direct Segment and Huge Pages
\subsection{TLB based approaches}
Efficient memory management, particularly in the context of 
Translation Lookaside Buffer (TLB) optimization, has been a focal point of 
research and development within computer architecture. Various techniques have been 
proposed to mitigate TLB-related bottlenecks and improve overall system performance.

\subsubsection{Huge Pages:}
This is used to map a very large region of memory to a 
single entry. This small/large region of memory is physically
contiguous. Most implementations of huge pages \cite{panwar_hawkeye_2019} are size
aligned, For example for the x86 architecture the huge pages 
size are 4KB, 2MB and 1GB pages. 
% \subsection{Tailored page sizes}
% TODO later
% \subsection{TLB coalescing}
% This leverages the default OS allocator behavior to pack
% multiple PTEs into a single TLB entry.

\subsubsection{Segment:}
A segment can be viewed as mapping between contiguous virtual
memory and contiguous physical memory. The property of a 
segment allows it to be larger than a page. Direct Segment allows the user to set a single segment
for an application. Two registers are added to mark the start
and end of the segment. Any virtual address within this region
can be translated by adding the fixed offset between the virtual
and physical address.

\subsubsection{Range Memory Mapping (RMM):}
RMM\cite{karakostas_redundant_2015} introduces the concept of adding an additional range table.
For large allocations RMM eagerly allocates contiguous physical pages.
The following allocations creates large memory ranges that are
both virtually and physically contiguous. RMM builds on the concept
of Direct segment by adding offset to translate a virtual address 
to physical address. RMM compares address with range boundaries 
to decide which range it belongs to. RMM queries the range table 
ofter an L1 TLB miss.

\subsubsection{FlexPointer:}
FlexPointer\cite{chen_flexpointer_2023} is based on the RMM\cite{karakostas_redundant_2015} paper. FlexPointer
does eagerly allocate pages which are physically contiguous and stores the ID to translate a virtual address 
to physical address on the remaining unused bits on the 64 bit virtual address. 
The paper contribution mentions shifting the TLB lookup to an earlier stage to improve
latency of accessing the TLB entries. FlexPointer immediate queries the 
range TLB for translations rather than the RMM paper which waits for the L1 TLB miss.

\subsection{CHERI}
A capability functions as a token that provides the holder with the authority to 
execute specific actions. By carefully managing who possesses these capabilities, 
it is possible to enforce security measures, such as ensuring that a particular
section of software can only read from a designated range of memory addresses.

The concept of capabilities has a rich history in computer science, tracing 
back to early systems designed to enhance security and manage resources effectively.
For instance, foundational work discussed in sources like [7] offers a comprehensive 
narrative on the evolution of capability architectures. This historical perspective 
can be further enriched by incorporating insights from more recent studies, 
such as those found in [11].

CHERI (Capability Hardware Enhanced RISC Instructions) represents a modern embodiment of this 
long-standing idea. It introduces more granular control over permissions, allowing for finer 
distinctions in what actions can be performed and by whom. Moreover, CHERI is designed to integrate 
seamlessly with contemporary processor instruction sets, ensuring that these advanced security 
features can be implemented on modern hardware platforms. This adaptation not only revitalizes 
the capability concept but also significantly enhances its applicability and effectiveness 
in current computing environments.

\section{FAT-Pointer based range addresses}
% This experiment exploits the FAT pointer to enable faster

% TODO: 
% - Add references for FlexPointer, Range Memory Mapping (RMM) and Direct segment
% \subsection{Overview}

FAT-Pointers, combined with the capabilities of the CHERI (Capability Hardware Enhanced RISC Instructions) 
architecture, introduce robust memory safety and security features by incorporating additional metadata 
with memory pointers. This enhanced architecture utilizes concepts such as FlexPointer, 
Range Memory Mapping (RMM) to manage memory effectively.

@@

% \begin{minipage}[t]{0.4\linewidth}
\begin{figure}[h]
  \includegraphics[width=0.8\textwidth]{diagrams/HighOverviewArchitecture24.png}
  \caption{High overview architecture}
  \label{fig:HighOverviewArchitecture}
% \end{minipage}
\end{figure}

% The figure above demonstrates the approach taken for using CHERI 
% 128bit FAT pointer scheme to allow a blocked based behavior on 
% on physically contiguous memory which is on the right against the 
% regular mmap approach which further elaborated in section (). The 
% green highlighted refers to the excess space available between 
% the 48th and 64th bit which can be used to store more meta-data which 
% is elaborated in the future work section().

Figure \ref{fig:HighOverviewArchitecture} illustrates the methodology employed to leverage the CHERI 
128-bit FAT-pointer scheme for facilitating block-based memory management
 on physically contiguous memory, which is depicted on the right side of the figure. 
 This technique contrasts with the conventional mmap approach, the details of which 
 are elaborated in section ("To be added later").

In figure \ref{fig:HighOverviewArchitecture}, the green-highlighted section marks the unused space between the 48th and 64th bits
within the FAT-pointer. This area of unused bits presents an opportunity to store additional metadata,
potentially enhancing the capabilities of the memory management system. The potential applications of 
this extra space are discussed in the future work section ("To be added later"), 
where we explore how this additional metadata storage could be used to further optimize memory allocation.

% By employing the CHERI 128-bit FAT-pointer scheme, the approach depicted aims to streamline the management of memory blocks, ensuring they are physically contiguous. This is in contrast to the traditional mmap approach, which typically does not guarantee physical contiguity of allocated memory regions.

\subsection{Range creation and huge pages}
% \begin{minipage}[t]{0.4\linewidth}
\begin{figure}[h]
  \includegraphics[width=0.8\textwidth]{diagrams/AllocationOverview24.png}
  \caption{Range of memory}
  \label{fig:RangeOfMemory}
\end{figure}
% \end{minipage}

% Ranges of memory are created based on bounds encoded to the FAT-Pointer based on the CHERI 128 bit 
% bounds compressed scheme. The Chuck between the specified upper and lower bounds are always physically
% contiguous. In the following implementation an arbitrary size huge page is initially allocated and within 
% this huge page custom size memory is allocated using a custom written mmap function which overwrites the existing block 
% based mmap function. Once memory is physically allocated using the custom mmap function call then bounds are set to it for tracking 
% the block of memory instead of using the TLB.

In this implementation, memory ranges are established using bounds encoded within the FAT-pointer, adhering 
to the CHERI 128-bit bounds compression scheme\cite{woodruff_cheri_2019}. The memory chunk defined by the upper and lower bounds is 
always physically contiguous. Initially, a huge page of arbitrary size is allocated. Within this huge page, 
custom-sized memory segments are allocated using a custom-designed mmap function, which overrides the existing 
block-based mmap function. Once the memory is physically allocated through this custom mmap function, bounds 
are set to track the memory block, eliminating the need for traditional TLB usage for this purpose. Traditional TLB usage 
involves maintaining numerous TLB entries, often supplemented by an L2 TLB and other hierarchical structures, 
to translate virtual addresses to physical addresses. This approach requires multiple entries to handle various 
memory segments, leading to increased overhead and complexity in address translation. Conversely, 
the current approach streamlines this process by using a single TLB entry to translate multiple
 addresses within a contiguous memory range. This reduces the number of required TLB entries, 
 simplifying the translation process and improving efficiency. By consolidating address translations 
 into a single TLB entry, this method minimizes the overhead associated with managing numerous TLB entries 
 and leverages the bounds encoded within the FAT-pointer for efficient memory tracking and access. 
 This approach allows for precise and efficient memory management within the allocated huge page.

%  The figure mentioned above demonstrates a simple use-case were the dark pink line refers to a huge page and 
%  the orange and blue lines are 2 equivalent to 2 Malloc calls allocating memory in different regions which in turn
%  emulates a block based memory allocator within a huge page using the bounds encoded in the FAT-Pointer.
\smallskip\noindent
Figure \ref{fig:RangeOfMemory} illustrates a straightforward use-case in which the dark pink line represents a single, 
large contiguous memory area, or huge page. Within this huge page, the orange and blue lines indicate 
two separate memory allocations equivalent to invoking malloc twice to allocate memory in distinct regions. 
This scenario simulates a block-based memory allocator operating within the confines of the huge page. 
The allocations leverage the bounds encoded in the FAT-pointer, ensuring tracking and efficient 
management of the allocated memory regions. By using the FAT-pointer bounds, this method maintains the 
integrity and contiguity of the allocated blocks within the huge page.

% This is done via the FreeBSD kernel Contiguous memory allocator.

% \subsection{Fragmentation}
% % The problem with standard allocators which are physically contiguous
% % is that they are mostly on the
% todo
% \subsection{Allocation with huge pages}

\subsection{Software Stack}
\begin{figure}[h]
% \begin{minipage}[t]{0.4\linewidth}
\includegraphics[width=0.8\textwidth]{diagrams/SoftwareStack24.png}
\caption{Overview of the software stack}
\label{fig:SoftwareStack}
% \end{minipage}
\end{figure}

% The Software stack consists of CHERIBSD as the base operating system due to official 
% support by ARM to the Morello's performance counters. As mentioned in the figure above 
% there is a C program that is linked to the prototype memory allocator or allocators benchmarked 
% against as mentioned in section (x) as either a Shared object file at compile time or as a header 
% file for smaller memory allocators. The modified mmap function call which is designed to be physically 
% contiguous is linked to the contigmem driver which is modified from the DPDK library (reference).
% The contigmem driver is loaded on boot time and reserves arbitrary size of a huge page which is
% set based on the experiment conducted.

The software stack is based on CHERIBSD, selected because ARM officially supports Morello's performance 
counters on this operating system. As illustrated in the figure \ref{fig:SoftwareStack}, the setup includes a C program that 
is linked to the prototype memory allocator or to various memory allocators being benchmarked, as described 
in section ("Evaluation section"). This linkage can occur in two ways: either as a shared object file during compile time 
for larger allocators, or as a header file for smaller allocators, ensuring flexibility and efficiency 
in memory management.

The custom mmap function, tailored to ensure physically contiguous memory allocation, is a key component 
of this system. This function is linked to the contigmem driver, which has been modified from the DPDK library 
to meet the specific needs of this implementation. The contigmem driver is essential for managing large contiguous 
memory blocks and is loaded during the system boot process. It reserves a huge page of arbitrary size, with the 
size parameter set based on the requirements of the conducted experiments.

@@

\section{Evaluation}
To evaluate the performance of FAT-Pointer based range addresses a sample implementation we used
the morello board with CheriBSD's Benchmark ABI compilation mode for accurate performance recordings. 
The evaluation is to identify CheriBSD's default memory allocator SnMalloc against the prototype memory 
allocator using the contribution of this paper in terms of: 
To assess the performance of FAT-Pointer-based range addressing in the sample implementation, we utilized the 
Morello board running CheriBSD in Benchmark ABI compilation mode to ensure precise performance measurements. 
The evaluation focuses on comparing CheriBSD's default memory allocator, SnMalloc, against the prototype 
memory allocator developed in this study, with particular attention to the following aspects:

\begin{table}[!ht]
  \centering
  \begin{tabular}{|l|l|l|l|l|}
  \hline
      Metric name & type of graph & tool used & x axis & y axis \\ \hline
      DTLB L1 read & line graph & Pmcstat & Time & DTLB L1 reads (each second) \\ \hline
      DTLB L2 read & line graph & Pmcstat & Time & DTLB L2 reads (each second) \\ \hline
      DTLB walk & line graph & Pmcstat & Time & DTLB Walks (each second) \\ \hline
      L1 cache miss & line graph & Pmcstat & Time & L1 cache miss (each second) \\ \hline
      Wall clock run time & bar graph & time & Benchmarks & Time \\ \hline
  \end{tabular}
\end{table}

\subsection{Benchmarks used}:
To conduct the evaluations, we utilized the COZ benchmark suite, a well-regarded tool specifically designed 
to measure and analyze performance improvements in concurrent programs. The COZ benchmark suite provides a
robust framework for identifying bottlenecks and evaluating the performance impact of various optimization
techniques. By leveraging COZ, developers can gain precise insights into the efficiency and scalability of
their concurrent code, making it an ideal choice for rigorous performance analysis.

From the extensive set of benchmarks provided by COZ, we selected four representative C programs. 
These programs were chosen based on their relevance to common concurrent programming patterns and
their ability to effectively demonstrate the strengths and weaknesses of different optimization 
strategies. The selected programs cover a range of concurrency scenarios, ensuring a comprehensive
evaluation of performance improvements.

By implementing these modifications, we ensured that the selected C programs not only adhered to
CHERI's security model but also maintained their functional and performance characteristics. 
This allowed us to effectively use the COZ benchmark suite to analyze the performance in a CHERI-enhanced environment.

\begin{figure}[h]
  \includegraphics[width=0.8\textwidth]{diagrams/expirement-runs.png}
\end{figure}

% \begin{itemize} % Coz C programs
%   \item Kmeans
%   \item Histogram
%   \item Matrix multiply
% \end{itemize}

\subsection{Future work}
% The future plan is to transition from the current experiment, which involves working on the ARM architecture on the ARM Morello board. 
% The current limitation is that all memory reads must go through the TLB for translations. The future plan involves storing 
% the offset directly on the pointer and using the bounds in CHERI to enable block-based allocator behavior. This phase of the 
% experiment is intended to be conducted on the RISC-V implementation of CHERI, known as Tooba.

>>>>

%%
%% This is file `sample-sigconf-authordraft.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,authordraft')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf-authordraft.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigconf,authordraft]{acmart}
\usepackage{listings}
\lstset{basicstyle=\small\ttfamily,columns=fullflexible}

@@

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation emai}{June 03--05,
  2018}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/18/06}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{FAT Pointer based range addresses}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Akilan Selvacoumar}
% \authornote{Both authors contributed equally to this research.}
% \email{as251@hw.ac.uk}
% \orcid{1234-5678-9012}
% \author{G.K.M. Tobin}
\authornotemark[1]
\email{as251@hw.ac.uk}
\affiliation{%
  \institution{Heriot Watt}
  \city{Edinburgh}
  \state{Scotland}
  \country{United Kingdom}
}

% \author{Lars Th{\o}rv{\"a}ld}
% \affiliation{%
%   \institution{The Th{\o}rv{\"a}ld Group}
%   \city{Hekla}
%   \country{Iceland}}
% \email{larst@affiliation.org}

@@

% \author{Aparna Patel}
% \affiliation{%
%  \institution{Rajiv Gandhi University}
%  \city{Doimukh}
%  \state{Arunachal Pradesh}
%  \country{India}}

% \author{Huifen Chan}
% \affiliation{%
%   \institution{Tsinghua University}
%   \city{Haidian Qu}
%   \state{Beijing Shi}
%   \country{China}}

% \author{Charles Palmer}
% \affiliation{%
%   \institution{Palmer Research Laboratories}
%   \city{San Antonio}
%   \state{Texas}
%   \country{USA}}
% \email{cpalmer@prl.com}

% \author{John Smith}
% \affiliation{%
%   \institution{The Th{\o}rv{\"a}ld Group}
%   \city{Hekla}
%   \country{Iceland}}
% \email{jsmith@affiliation.org}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
The increasing disparity between application workloads and the capacity of Translation Lookaside Buffers (TLB) 
has prompted researchers to explore innovative solutions to mitigate this gap. One such approach involves 
leveraging physically contiguous memory to optimize TLB utilization. Concurrently, advancements in hardware-level 
system security, exemplified by the Capability Hardware Enhanced RISC Instructions (CHERI) architecture, offer 
additional opportunities for improving memory management and security.

@@

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
% %%
% \begin{CCSXML}
% <ccs2012>
%  <concept>
%   <concept_id>00000000.0000000.0000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>500</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>00000000.00000000.00000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>300</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>00000000.00000000.00000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>00000000.00000000.00000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
% </ccs2012>
% \end{CCSXML}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
% \keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
%   Your, Paper}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
% \begin{teaserfigure}
%   \includegraphics[width=\textwidth]{sampleteaser}
%   \caption{Seattle Mariners at Spring Training, 2010.}
%   \Description{Enjoying the baseball game from the third-base
%   seats. Ichiro Suzuki preparing to bat.}
%   \label{fig:teaser}
% \end{teaserfigure}

@@

@@

\section{Introduction}
In the dynamic landscape of computing, the pursuit of optimal performance is a constant endeavor, 
especially as applications evolve to handle increasingly complex workloads. 
One critical aspect influencing performance is memory management, where efficient 
utilization of resources is paramount. Translation Lookaside Buffers (TLBs) play a 
pivotal role in this regard, expediting memory access by storing recently accessed memory translations.
However, as applications grow in size and complexity, the capacity of TLBs often struggles to 
keep pace, leading to performance bottlenecks\cite{mittal_survey_2017}. To address this challenge, researchers have 
turned to innovative solutions, one of which involves harnessing the benefits of huge pages.
Huge pages, also known as large pages, allow for the allocation of memory in significantly 
larger chunks compared to traditional small pages. By reducing the number of TLB entries 
needed to access a given amount of memory, huge pages offer a potential avenue for optimizing 
TLB utilization and thereby enhancing overall system performance.

@@

In this context, the integration of huge pages into memory management 
strategies alongside capability-based addressing in architectures like 
CHERI offers a compelling synergy. By optimizing TLB utilization through the 
utilization of huge pages and leveraging the security features of capability-based addressing, 
significant performance improvements can be realized. This approach not only enhances 
system security but also accelerates memory access. The following below are research questions 
we are addressing:
\newline

\begin{enumerate}
    \item How does the utilization of bounds for tracking memory allocations, in addition to security purposes, affect the 
    run times and Translation Lookaside Buffer (TLB) miss rates in modern computing systems?
    \item How does the implementation of bounds for seeking through physically contiguous memory influence the complexity and 
    efficiency of standard memory allocators, particularly those with advanced features such as transparent 
    huge pages, and what are the implications for system performance in terms of execution speed, memory access 
    latency, and resource utilization?
\end{enumerate}

\section{FAT Pointer based range addresses}
FAT-Pointers based range addresses, combined with the capabilities of the CHERI (Capability Hardware Enhanced RISC Instructions) 
architecture, introduce robust memory safety and security features by incorporating additional metadata 
with memory pointers. This enhanced architecture utilizes concepts such as FlexPointer, 
Range Memory Mapping (RMM) to manage memory effectively.

@@

\begin{figure}[h]
  \includegraphics[width=0.4\textwidth]{diagram/HighOverviewArchitecture24.png}
  \caption{High overview architecture}
  \label{fig:HighOverviewArchitecture}
% \end{minipage}
\end{figure}

Figure \ref{fig:HighOverviewArchitecture} illustrates the methodology employed to leverage the CHERI 
128-bit FAT-pointer scheme for facilitating block-based memory management
on physically contiguous memory, which is depicted on the right side of the figure. 
This technique contrasts with the conventional mmap approach.

In figure \ref{fig:HighOverviewArchitecture}, the green-highlighted section marks the unused space between the 48th and 64th bits
within the FAT-pointer. This area of unused bits presents an opportunity to store additional metadata,
potentially enhancing the capabilities of the memory management system. 
Here we explore how this additional metadata storage could be used to further optimize memory allocation.

\subsection{Range creation and huge pages}
\begin{figure}[h]
  \includegraphics[width=0.4\textwidth]{diagram/AllocationOverview24.png}
  \caption{Range of memory}
  \label{fig:RangeOfMemory}
\end{figure}
In this implementation, memory ranges are established using bounds encoded within the FAT-pointer, adhering 
to the CHERI 128-bit bounds compression scheme\cite{woodruffcheri2019}. The memory chunk defined by the upper and lower bounds is 
always physically contiguous. Initially, a huge page of arbitrary size is allocated. Within this huge page, 
custom-sized memory segments are allocated using a custom-designed mmap function, which overrides the existing 
block-based mmap function. Once the memory is physically allocated through this custom mmap function, bounds 
are set to track the memory block, eliminating the need for traditional TLB usage for this purpose. Traditional TLB usage 
involves maintaining numerous TLB entries, often supplemented by an L2 TLB and other hierarchical structures, 
to translate virtual addresses to physical addresses. This approach requires multiple entries to handle various 
memory segments, leading to increased overhead and complexity in address translation. Conversely, 
the current approach streamlines this process by using a single TLB entry to translate multiple
addresses within a contiguous memory range. This reduces the number of required TLB entries, 
simplifying the translation process and improving efficiency. By consolidating address translations 
into a single TLB entry, this method minimizes the overhead associated with managing numerous TLB entries 
and leverages the bounds encoded within the FAT-pointer for efficient memory tracking and access. 
This approach allows for precise and efficient memory management within the allocated huge page.

\smallskip\noindent
Figure \ref{fig:RangeOfMemory} illustrates a straightforward use-case in which the dark pink line represents a single, 
large contiguous memory area, or huge page. Within this huge page, the orange and blue lines indicate 
two separate memory allocations equivalent to invoking malloc twice to allocate memory in distinct regions. 
This scenario simulates a block-based memory allocator operating within the confines of the huge page. 
The allocations leverage the bounds encoded in the FAT-pointer, ensuring tracking and efficient 
management of the allocated memory regions. By using the FAT-pointer bounds, this method maintains the 
integrity and contiguity of the allocated blocks within the huge page.

\subsection{Software Stack}
The software stack is based on CHERIBSD, selected because ARM officially supports Morello's performance 
counters on this operating system. As illustrated in the figure \ref{fig:SoftwareStack}, the setup includes a C program that 
is linked to the prototype memory allocator or to various memory allocators being benchmarked. This linkage can occur in two ways: either as a shared object file during compile time 
for larger allocators, or as a header file for smaller allocators, ensuring flexibility and efficiency 
in memory management.
\newline

@@

\subsection{Contigmem driver from DPDK}
The custom mmap function, tailored to ensure physically contiguous memory allocation, is a key component 
of this system. This function is linked to the contigmem driver, which has been modified from the DPDK\cite{bidpdk-based2016} library 
to meet the specific needs of this implementation. The contigmem driver is essential for managing large contiguous 
memory blocks and is loaded during the system boot process. It reserves a huge page of arbitrary size, with the 
size parameter set based on the requirements of the conducted experiments.

\begin{lstlisting}[language=C, caption=Contigmem driver , label=ContigInit]

  MALLOC_DEFINE(M_CONTIGMEM, "contigmem", 
  "contigmem(4) allocations");

  static int contigmem_modevent(module_t mod, 
  int type, void *arg)
  {
    int error = 0;

    switch (type) {
    case MOD_LOAD:
      error = contigmem_load();
      break;
    case MOD_UNLOAD:
      error = contigmem_unload();
      break;
    default:
      break;
    }

    return error;
  }

  ....

DECLARE_MODULE(contigmem, contigmem_mod, 
SI_SUB_DRIVERS, SI_ORDER_ANY);
MODULE_VERSION(contigmem, 1);

static struct cdevsw contigmem_ops = {
	.d_name         = "contigmem",
	.d_version      = D_VERSION,
	.d_flags        = D_TRACKCLOSE,
	.d_mmap_single  = contigmem_mmap_single,
	.d_open         = contigmem_open,
	.d_close        = contigmem_close,
};

static int
contigmem_load()
{
	....

	for (i = 0; i < contigmem_num_buffers; i++) {
		addr = contigmalloc(contigmem_buffer_size, 
           M_CONTIGMEM, M_ZERO,
		   0, BUS_SPACE_MAXADDR, 
           contigmem_buffer_size, 0);
	....
	}

    ....

error:
	for (i = 0; i < contigmem_num_buffers; i++) {
		if (contigmem_buffers[i].addr != NULL) {
			contigfree(contigmem_buffers[i].addr,
				contigmem_buffer_size, M_CONTIGMEM);
			contigmem_buffers[i].addr = NULL;
		}
		....
	}

	return error;
}

\end{lstlisting}

When the contigmem\_load function is called, either during boot or when the Kernel module is loaded, it pre-allocates 
a segment of physically contiguous memory. This approach differs from FlexPointer\cite{chenflexpointer2023}, 
which allocates physically contiguous memory eagerly. The contigmem\_load function allocates memory using contigmalloc, 
which allocates physically contiguous memory initialized to zero. The cdevsw struct refers to function 
calls which would be overwritten on loading the driver. In the code snippet above
the mmap function would be overwritten with contigmem\_mmap\_single if the following driver is opened and truncated
as shown in Code snippet. 
\newline

In the code snippet the cdev\_pager\_ops refers to the operations which will be overwritten 
when called with mmap such as overwriting page faults.

% \begin{lstlisting}[language=C, caption=Contigmem driver mmap , label=ContigMmap]
